Computing exact restricted isometry constants via\\ mixed integer semidefinite programming

### 2016-03-11 RIP-MISDP-Paper/IEEE_RIP-MISDP.tex
reworked introduction

One important condition is the \emph{restricted isometry property} (RIP),
which was first introduced by Cand\`es and Tao~\cite{CT05}. For each $k \in
\{1, \dots, n\}$, the RIP involves a \emph{restricted isometry constant}
$\delta_k$ bounding the $k$-sparse eigenvalues of $A$. If $\delta_k$ is sufficiently small
and there exists $\tilde{x} \in \R^n$ with $A \tilde{x} = b$ and
$\Norm{\tilde{x}}{0} \leq k$, then problems~\eqref{l0min} and~\eqref{l1min}
are equivalent; see Section~\ref{sec:RIP} for more details. There are
several types of matrices~$A$ for which the RIP holds, for instance random
matrices, see, e.g., Baraniuk et al.~\cite{BDDW08}. Although the RIP does
not provide necessary conditions, it is one of the major tools for
establishing the $\ell_0$-$\ell_1$-equivalence.

We note that recovery results can also be derived for the \emph{denoising}
case, in which instead of~\eqref{l0min} one considers
\begin{align}\label{l0minDenoising}
  \min \quad & \Norm{x}{0} \tag{$P_0^\delta$}\\
  \text{s.t.} \quad & \Norm{Ax - b}{2} \leq \delta, \nonumber
\end{align}
for some $\delta \geq 0$; the $\ell_1$-problem~\eqref{l1min} is changed
similarly. For more details we again refer the reader to~\cite{FouR13}.

However, one drawback of the RIP is that computing the restricted isometry
constants for a given matrix~$A$ and~$k$ is NP-hard, see~\cite{PT14}. On
the other hand, d'Aspremont et al.~\cite{Asp08,Asp07} proposed several
semidefinite relaxations to compute upper bounds on the restricted isometry
constants. Note that small upper bounds suffice to guarantee
$\ell_0$-$\ell_1$-equivalence between~\eqref{l0min} and~\eqref{l1min}, but
there exist cases in which the bounds are too weak to give a guarantee.

In the following, we will start by introducing the restricted isometry
property and corresponding results about the equivalence of~\eqref{l0min}
and~\eqref{l1min} in Section~\ref{sec:RIP}. In Section~\ref{sec:SDPrelax},
we will then review the SDP-relaxations of~\cite{Asp08}
and~\cite{Asp07}. Afterwards we will construct a mixed integer semidefinite
programm and show that it produces exact restricted isometry constants in
Section~\ref{sec:MISDP}. Finally, in Section~\ref{sec:numerical_results},
we present numerical results and comparisons with the SDP relaxations in
\cite{Asp08} and \cite{Asp07}.

In this section we will review some general results about the restricted
isometry property and its relationship with the optimization problems
(\ref{l0min}) and (\ref{l1min}). We start with the definition of the RIP as
in \cite{CT05}:

@@

As the following results depend on the constant $\delta$ in (\ref{rip}) being
small enough, we are generally interested in the smallest constant satisfying
(\ref{rip}).

@@

@@

 % If the restricted isometry property holds for a sufficiently small restricted isometry constant, then the optimal solutions of (\ref{l0min}) and (\ref{l1min}) coincide. To show this, one needs estimates of the form
 %  \begin{equation}\label{l1ErrorRIP}
 %  \Norm{x^* - x} \leq C \Norm{x - x_k},
 % \end{equation}
 % where $x^*$ is an optimal solution of (\ref{l1min}) with $b = Ax$ and $x_k$ is the best approximation of some vector $x \in \R^n$ with $k$-sparse vectors. If we get a result like (\ref{l1ErrorRIP}), and we know that the vector $x$, which in the 
 % applications would be the sparse signal that needs to be recovered, is already $k$-sparse itself, then the best approximation of $x$ with $k$-sparse vectors is clearly $x$ itself. This means that the right-hand-side in 
 % (\ref{l1ErrorRIP}) equals zero, and therefore the unique optimal solution of (\ref{l1min}) has to be $x$ again. Now (\ref{l1ErrorRIP}) not only holds for this $x$ exclusively, but for any $k$-sparse vector $y$ with $Ay=Ax=b$. And 
 % because of $Ay=Ax=b$, we get the same problem in (\ref{l1min}) and therefore also the same solution $x^*$ which needs to equal $y$ because of (\ref{l1ErrorRIP}), so this means that there can only be a single vector $x$ with sparsity
 % at most $k$ for which $Ax=b$ holds, which then also has to solve (\ref{l0min}), which gives us the equivalance of these two problems.

There are also many other results similar to Theorem~\ref{RIPsqrt2}: For
example, Foucart~\cite{fou10} relaxed the condition to $\delta_{2k} <
0.4652$ and Foucart and Rauhut~\cite{FouR13} to $\delta_{2k} < 0.6246$.
\comment{Ergebnis ist f\"ur $\C$, sollte aber auch f\"ur $\R$
gelten, da man RIC \"uber $\C$ durch Aufspaltung in Real- und 
Imagin\"arteil durch reelle RIC absch\"atzen k\"onnen sollte.}
Moreover, Cai et al.~\cite{CWX10} gave the condition $\delta_k <
0.307$ for Theorem~\ref{RIPsqrt2} to hold. Note that all of these results
can be generalized to the denoising case~\eqref{l0minDenoising}.

Similar results hold for the asymmetric case. For instance, we have:

@@

Again this result can be generalized to the denoising case. One advantage
of Definition~\ref{asymRIPdef} and Theorem~\ref{asymRIPtheorem} is that they
are invariant under scaling of $Ax = b$. The condition of
Theorem~\ref{RIPsqrt2} holding for~$A x = b$ does not imply it for the scaled
$\lambda Ax = \lambda b$. However, in Definition \ref{asymRIPdef}, we can
use $\tilde{\alpha}_k = \lambda \alpha_k$, $\tilde{\beta}_k = \lambda
\beta_k$, and therefore $\tilde{\gamma}_k = \lambda^2 \alpha_k^2 /
\lambda^2 \beta_k^2 = \gamma_k$.

In the following we will design an approach to compute the optimal
constants $\alpha_k$ and $\beta_k$. These can then be used to compute
either the restricted isometry ratio $\gamma_k \define \beta_k^2 / \alpha_k^2$
or the restricted isometry constant $\delta_k = \max \{ 1 - \alpha_k^2,\;
\beta_k^2 - 1 \}$.

@@

In this section we will discuss two known SDP relaxations to compute an upper bound on the restricted isometry constant $\delta_k$ in (\ref{rip}), as they have some ideas in common with the mixed integer semidefinite program we want to 
propose and will also be checked against in the section on numerical results. The idea of these two relaxations as well as the MISDP in the next section is that the optimal constant $\alpha_k^2$ in (\ref{lowerRIP}) can be computed
by the non-convex quadratic optimization problem
\begin{align}\label{QP}
 \min \quad & \Norm{Ax}{2}^2 \nonumber \\
 \text{s.t.} \quad & \Norm{x}{2}^2 = 1 \tag{QP} \\
 & \Norm{x}{0} \leq k. \nonumber
\end{align}
The corresponding maximization problem can be solved to compute $\beta_k^2$. In \cite{Asp07} the non-convex quadratic equality constraint $\Norm{x}{2}^2 = 1$ is tackled by the technique of semidefinite lifting first introduced in \cite{GW95} 
for the max-cut problem. The idea is to use a new matrix variable $X=xx\T$ for the quadratic terms, where the condition $X=xx\T$ is then enforced by the equivalent constraints $X \succeq 0$ and $\Rk(X) = 1$, where 
$X \succeq 0$ means that $X$ is symmetric (shortly written \mbox{$X \in S_n$}) and positive semidefinite. The objective and the normalization-constraint can then be expressed in $X$ as  $\Norm{Ax}{2}^2 = x\T A\T Ax =
\Tr(A\T A X)$ and $\Norm{x}{2}^2 = \sum_{i=1}^n x_i^2 = \Tr(X)$. In the
next step the inequality $\Norm{x}{0} \leq k$ is then substituted by the coarser constraint $\ones\T
\abs{X} \ones \leq k$, which follows from the fact that $\ones\T\abs{X} \ones = \Norm{\vect(X)}{1}\leq \sqrt{\Norm{\vect(X)}{0}} \Norm{\vect(X)}{2}$ and $\Norm{\vect(X)}{0}=\Norm{x}{0}^2$ in the rank 1 case, where $\vect(X)$ is a vector
in $\R^{n^2}$ consisting of all entries of $X$. 
Relaxing the non-convex Rank-constraint finally leads to the following relaxation:
\begin{align}\label{Asp07}
 \min \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & \ones\T \abs{X} \ones \leq k \tag{A1} \\
 & X \succeq 0. \nonumber
\end{align}

@@

The testset consists of 63 matrices of seven 
different types including band matrices 
with band size three or five with entries within the band chosen uniformly in $\{0,1\}$, binary matrices with all entries chosen 
uniformly in $\{0,1\}$, Gaussian-distributed matrices and rank one 
matrices $A = aa\T$ with $\mathcal{N}(0,1)$-distributed entries of $a$. In addition to these four kinds of matrices, 
we also use three more with theoretical results about their restricted isometry constants, namely matrices with 
$N(0, 1/m)$-distributed entries, with entries chosen uniformly in $\pm 1/\sqrt{m}$ and finaly with distribution
\begin{equation}\nonumber
A_{ij} = \begin{cases} + \sqrt{3/m} & \text{with probability } \frac{1}{6} \\ 0 & \text{with probability } \frac{2}{3} \\ - \sqrt{3/m} & \text{with probability } \frac{1}{6}. \end{cases}
\end{equation}
For the latter three kinds of matrices it is shown in \cite{BDDW08} that for $n \rightarrow \infty$, 
sufficiently small $k$ and given $\delta$, they satisfy the restricted isometry property (\ref{rip}) of order $k$  
with probability converging exponentially to one. The needed proportion between $n$ and $k$ is so small, however,
that we cannot expect the generated matrices in small dimensions to satisfy the RIP. For each type we used \mbox{MATLAB 8.3.0} 
to randomly generated nine matrices of three different sizes with rows ranging from 15 to 30, columns 
from 25 to 40 and order $k$ between three and five.

@@

In Table~\ref{lhsGap} we compare the relative gap between the bounds on $\alpha_k$ computed by the two heuristics
and the exact values given by our
mixed integer semidefinite program. We give arithmetic means over all of the nine matrices for the given type that
could be solved within the time- and memory-limit. Note that we ommitted the rank one matrices, since these satisfy
$\alpha_k = 0$ in all cases, so that the trivial lower bound would already be optimal. For the 
left-hand-side of the RIP, both relaxations did not perform
too well. Especially the relaxation given in \cite{Asp08} regularly returned worse solution than the
trivial lower bound of zero. The relaxation in \cite{Asp07} performed slightly better, but the bounds
were still not nearly good enough to have any chance of proving $\ell_0$-$\ell_1$-equivalance. The relatively
bad performance of both relaxations for the left-hand side might not be too surprising, since both relaxations
were originally designed for sparse principal component analysis, which corresponds to the right-hand side
of the RIP.

@@

Results for the right-hand side, i.e. $\beta_k$, are given in Table~\ref{rhsGap}. Here both relaxations
perform much better. Problem~\eqref{Asp08Dual} constantly produced bounds within 10 to 35\,\% of the optimal
solution. The formulation in~\eqref{Asp08Dual} showed bigger fluctuation, but was very tight in many cases,
with a gap of less than two percent for band matrices.

@@

Solving times for the left- and right-hand side are given in Table~\ref{lhsTime} and~\ref{rhsTime}. The relaxation proposed in \cite{Asp07}
was very fast, taking only a few seconds for all instances. The mixed-integer semidefinite program obviously
cannot compete with these solving times, although for band-matrices, specifically, the difference was relatively
small. The Lagrangian Relaxation from \cite{Asp08}, however, took more time to produce bounds than was needed
by our MISDP to compute exact solutions. The big difference is also caused by the instances running into the 
memory limit, but even without those, the algorithm still took slightly longer than the MISDP.

@@

We have seen that exact restricted isometry constants can be computed not only via total enumeration but also by solving a mixed integer semidefinite program. Because of this exactness result, the new problem could drastically
improve the bounds computed with the semidefinite relaxations proposed in \cite{Asp07} and \cite{Asp08}, but this came at the cost of a much higher complexity of computation, at least in comparison to \cite{Asp07}. We could only 
solve these problems for relatively small instances, but this could change with sufficient advances in the area of mixed integer semidefinite programming, especially with the addition of special cutting planes, which proved to be 
the main enhancement needed to make mixed integer linear solvers as fast as they are today. Also special heuristics and branching rules could improve the performance of the branch-and-bound-algorithm and allow us to solve bigger 
instances of the proposed MISDP.

>>>>

One important condition is the \emph{restricted isometry property} (RIP),
which was first introduced by Cand\`es and Tao~\cite{CT05}.

@@

Since the following results depend on the constant $\delta$ in (\ref{rip}) being
small enough, we are generally interested in the smallest constant satisfying
(\ref{rip}).

@@

There are also many other results similar to Theorem~\ref{RIPsqrt2}: For
example, Foucart~\cite{fou10} relaxed the condition to $\delta_{2k} <
0.4652$ and Foucart and Rauhut~\cite{FouR13} to $\delta_{2k} < 0.6246$.
Moreover, Cai et al.~\cite{CWX10} gave the condition $\delta_k <
0.307$ for Theorem~\ref{RIPsqrt2} to hold.

@@

For the asmmetric version of the RIP in Definition~\ref{asymRIPdef}, we
can again show $\ell_0$-$\ell_1$-equivalence:

@@

One advantage
of Definition~\ref{asymRIPdef} and Theorem~\ref{asymRIPtheorem} is that they
are invariant under scaling of $Ax = b$. The condition of
Theorem~\ref{RIPsqrt2} holding for~$A x = b$ does not imply the same 
for the scaled equation
$\lambda Ax = \lambda b$. In Definition \ref{asymRIPdef}, however, we can
use $\tilde{\alpha}_k = \lambda \alpha_k$, $\tilde{\beta}_k = \lambda
\beta_k$, and therefore $\tilde{\gamma}_k = \lambda^2 \alpha_k^2 /
\lambda^2 \beta_k^2 = \gamma_k$.

Furthermore, the recovery results given in this section can also be derived 
for the \emph{denoising} case, where instead of~\eqref{l0min} one considers
\begin{align}\label{l0minDenoising}
  \min \quad & \Norm{x}{0} \tag{$P_0^\delta$}\\
  \text{s.t.} \quad & \Norm{Ax - b}{2} \leq \delta, \nonumber
\end{align}
for some $\delta \geq 0$; the $\ell_1$-problem~\eqref{l1min} is changed
similarly. For more details we refer the reader to~\cite{FouR13}.

\subsection{Literature Overview and Contribution of this Work}

There are several types of matrices~$A$ for which it is possible
to show analytically that the RIP holds, for instance random
matrices, see, e.g., Baraniuk et al.~\cite{BDDW08}.
Still, a big drawback of the RIP is that computing the restricted isometry
constants for a given matrix~$A$ and~$k$ is NP-hard, see~\cite{PT14}. 
On the other hand, d'Aspremont et al.~\cite{Asp08,Asp07} proposed several
semidefinite relaxations to compute upper bounds on the restricted isometry
constants. Note that small upper bounds suffice to guarantee
$\ell_0$-$\ell_1$-equivalence between~\eqref{l0min} and~\eqref{l1min}, but
there exist cases in which the bounds are too weak to give a guarantee.

In the following, we will start by reviewing the SDP-relaxations of~\cite{Asp08}
and~\cite{Asp07} in Section~\ref{sec:SDPrelax}.
Afterwards, we will construct a mixed integer semidefinite
programm and show that it produces exact restricted isometry constants in
Section~\ref{sec:MISDP}. Finally, in Section~\ref{sec:numerical_results},
we present numerical results and comparisons with the SDP relaxations in
\cite{Asp08} and \cite{Asp07}.

@@

In this section, we will discuss two known SDP relaxations to compute an upper bound on the restricted isometry constant $\delta_k$ in (\ref{rip}), as they have some ideas in common with the mixed integer semidefinite program we want to 
propose and will also be checked against in the section on numerical results. The idea of these two relaxations as well as the MISDP in the next section is that the optimal constant $\alpha_k^2$ in (\ref{lowerRIP}) can be computed
by the non-convex quadratic optimization problem
\begin{align}\label{QP}
 \min \quad & \Norm{Ax}{2}^2 \nonumber \\
 \text{s.t.} \quad & \Norm{x}{2}^2 = 1 \tag{QP} \\
 & \Norm{x}{0} \leq k. \nonumber
\end{align}
The corresponding maximization problem can be solved to compute $\beta_k^2$. In \cite{Asp07} the non-convex quadratic equality constraint $\Norm{x}{2}^2 = 1$ is tackled by the technique of semidefinite lifting first introduced in \cite{GW95} 
for the max-cut problem. The idea is to use a new matrix variable $X=xx\T$ for the quadratic terms, where the condition $X=xx\T$ is then enforced by the equivalent constraints $X \succeq 0$ and $\Rk(X) = 1$, where 
$X \succeq 0$ means that $X$ is symmetric (shortly written \mbox{$X \in S_n$}) and positive semidefinite. The objective and the normalization-constraint can then be expressed in $X$ as  $\Norm{Ax}{2}^2 = x\T A\T Ax =
\Tr(A\T A X)$ and $\Norm{x}{2}^2 = \sum_{i=1}^n x_i^2 = \Tr(X)$. In the
next step the inequality $\Norm{x}{0} \leq k$ is then substituted by the coarser constraint $\ones\T
\abs{X} \ones \leq k$, which follows from the fact that $\ones\T\abs{X} \ones = \Norm{\vect(X)}{1}\leq \sqrt{\Norm{\vect(X)}{0}} \Norm{\vect(X)}{2}$ and $\Norm{\vect(X)}{0}=\Norm{x}{0}^2$ in the rank 1 case, where $\vect(X)$ is a vector
in $\R^{n^2}$ consisting of all entries of $X$. 
Relaxing the non-convex Rank-constraint finally leads to the following relaxation:
\begin{align}\label{Asp07}
 \min \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & \ones\T \abs{X} \ones \leq k \tag{A1} \\
 & X \succeq 0. \nonumber
\end{align}

@@

The testset consists of 63 matrices of seven 
different types including band matrices 
with band size three or five with entries within the band chosen uniformly in $\{0,1\}$, binary matrices with all entries chosen 
uniformly in $\{0,1\}$, Gaussian-distributed matrices and rank one 
matrices $A = aa\T$ with $\mathcal{N}(0,1)$-distributed entries of $a$. In addition to these four kinds of matrices, 
we also use three more with theoretical results about their restricted isometry constants, namely matrices with 
$N(0, 1/m)$-distributed entries, with entries chosen uniformly in $\pm 1/\sqrt{m}$ and finally with distribution
\begin{equation}\nonumber
A_{ij} = \begin{cases} + \sqrt{3/m} & \text{with probability } \frac{1}{6} \\ 0 & \text{with probability } \frac{2}{3} \\ - \sqrt{3/m} & \text{with probability } \frac{1}{6}. \end{cases}
\end{equation}
For the latter three kinds of matrices it is shown in \cite{BDDW08} that for $n \rightarrow \infty$, 
sufficiently small $k$ and given $\delta$, they satisfy the restricted isometry property (\ref{rip}) of order $k$  
with probability converging exponentially to one. The needed proportion between $n$ and $k$ is so small, however,
that we cannot expect the generated matrices in small dimensions to satisfy the RIP. For each type we used \mbox{MATLAB 8.3.0} 
to randomly generate nine matrices of three different sizes with rows ranging from 15 to 30, columns 
from 25 to 40 and order $k$ between three and five.

@@

In Table~\ref{lhsGap} we compare the relative gap between the bounds on $\alpha_k$ computed by the two heuristics
and the exact values given by our
mixed integer semidefinite program. We give arithmetic means over all of the nine matrices for the given type that
could be solved within the time- and memory-limit. Note that we ommitted the rank one matrices, since these satisfy
$\alpha_k = 0$ in all cases, so that the trivial lower bound would already be optimal. For the 
left-hand-side of the RIP, both relaxations did not perform
too well. Especially the relaxation given in \cite{Asp08} regularly returned worse solutions than the
trivial lower bound of zero. The relaxation in \cite{Asp07} performed slightly better, but the bounds
were still not nearly good enough to have any chance of proving $\ell_0$-$\ell_1$-equivalance. The relatively
bad performance of both relaxations for the left-hand side might not be too surprising, since both relaxations
were originally designed for sparse principal component analysis, which corresponds to the right-hand side
of the RIP.

@@

Results for the right-hand side, i.e. $\beta_k$, are given in Table~\ref{rhsGap}. Here both relaxations
perform much better. Problem~\eqref{Asp08Dual} constantly produces bounds within 10 to 35\,\% of the optimal
solution. The formulation in~\eqref{Asp08Dual} shows bigger fluctuation, but is very tight in many cases,
with a gap of less than two percent for band matrices.

@@

Solving times for the left- and right-hand side are given in Table~\ref{lhsTime} and~\ref{rhsTime}. The relaxation proposed in \cite{Asp07}
is very fast, taking only a few seconds for all instances. The mixed-integer semidefinite program obviously
cannot compete with these solving times, although for band-matrices, specifically, the difference is relatively
small. The Lagrangian Relaxation from \cite{Asp08}, however, takes more time to produce bounds than is needed
by our MISDP to compute exact solutions. The big difference is also caused by the instances running into the 
memory limit, but even without those, the algorithm still takes slightly longer than the MISDP.

@@

We have seen that exact restricted isometry constants can be computed not only via total enumeration but also by solving a mixed integer semidefinite program. Because of this exactness result, the new problem can drastically
improve the bounds computed with the semidefinite relaxations proposed in \cite{Asp08} and \cite{Asp07}, but this comes at the cost of a much higher complexity of computation, at least in comparison to \cite{Asp07}. We could only 
solve these problems for relatively small instances, but this might change with sufficient advances in the area of mixed integer semidefinite programming, especially with the addition of special cutting planes, which proved to be 
the main enhancement needed to make mixed integer linear solvers as fast as they are today. Also special heuristics and branching rules could improve the performance of the branch-and-bound-algorithm and allow us to solve bigger 
instances of the proposed MISDP.
